---
title: "Data Overview and Preparation"
author: "Catalina Valencia"
output: html_document
---

**Project:** Credit Approval Classification: Model Validation and Generalization Analysis

**File:** 01_data_overview_and_preparation.Rmd

### Objectives:
- Load and inspect the Credit Approval dataset.
- Verify structure, variable roles, and completeness.
- Make pre-processing decisions (scaling, response definition).
- Create clean objects for downstream modeling.

### Notes:
- No modeling is performed in this file.

Environment Setup

```{r}
rm(list = ls())
```

Load Data

```{r}
ccdata <- read.table(
  "../data/credit_card_data.txt",
  header = FALSE,
  stringsAsFactors = FALSE
)
```

### Data Overview

```{r}
head(ccdata)
tail(ccdata)
dim(ccdata)
str(ccdata)
summary(ccdata)
```

**Results:**

- The dataset contains 654 observations and 11 variables: 10 predictors (V1–V10) and 1 binary response (V11).

- Predictors (V1–V10) are all numeric, consisting of a mix of continuous variables, integers, and binary variables (0/1).

- The response variable (V11) is a binary outcome representing credit approval status (0/1).

- The documentation does not explicitly define which value corresponds to approval or rejection; therefore, models treat class labels abstractly and are evaluated based on correct classification only.

The presence of predictors on different numeric scales supports the use of standardized features in downstream SVM and KNN models.

```{r}
table(ccdata$V11)
prop.table(table(ccdata$V11))
```

**Results:**

- The response variable V11 is binary, with moderately balanced classes.

- No severe class imbalance is present, so accuracy is an appropriate performance metric in subsequent analyses.

### Response and Predictor Definition

**Response variable:**

V11 indicates credit approval status (binary outcome)

```{r}
y <- as.factor(ccdata$V11)
```

**Predictor variables:**

V1 through V10 are numeric predictors

```{r}
X <- ccdata[, 1:10]
```

### Missing Data Verification

Confirm that the dataset contains no missing values

```{r}
colSums(is.na(ccdata))
```

**Results:**

- No missing values are present in the original dataset.

- Imputation experiments are conducted separately and do not affect the modeling dataset used in subsequent analyses.

### Scaling Decision

Predictors are scaled (centered and standardized).

Scaling is required for:

- Support Vector Machine Modeling (margin-based).
- k-Nearest Neighbor Modeling (distance-based).

```{r}
X_scaled <- scale(X)
```

Combine scaled predictors with response

```{r}
ccdata_scaled <- data.frame(X_scaled, V11 = y)
```

**Results:**

Predictors are standardized to ensure that all variables contribute comparably to distance-based (KNN) and margin-based (SVM) classifiers.

### Objects Created for Reference

ccdata        : original unscaled dataset 

y             : factor-encoded response variable 

X             : raw predictor matrix

X_scaled      : standardized predictors (demonstration only)

ccdata_scaled : combined scaled predictors and response

**Note:**

Downstream modeling scripts reload the dataset directly and perform scaling internally to ensure self-contained analyses.